\personalAcknowledgementEnglish{
First and foremost, I would like to thank the Dean and Professor Idit Keidar for being my Master's Thesis Supervisor. It was an honor and a privilege to work under your guidance. Thank you for believing in me and giving me the chance to learn from you, it has been an incredible and invaluable journey. 
As such, I am equally grateful to my fellow colleague and researcher Arik Rinberg. Thank you for being the constant anchor, to whom I could always turn to for ideas when I was stuck or simply to converse and learn from, your help made all the difference.
To the members of Professor Idit’s research group: Gal Assa, Oded Naor, Shir Cohen, and Ramy Fakhoury – your time, presence, insight, and constant support were indispensable, thank you for having my back. 
I would like to thank my family and friends for their love and continuous support. Especially, I would like to thank my husband Sagi for the endless amount of support, love, and encouragement to complete my academic journey. Thank you for always showing interest in my work and helping me get the best out of myself. 
Lastly, I couldn't have done this without my parents. I am forever grateful for their constant support, love, and belief in me.

}

\personalAcknowledgementHebrew{

בראש ובראשונה, אני רוצה להודות לדיקן ולפרופסור עידית קידר, המנחה שלי לתזה לתואר שני. זה היה כבוד וזכות לעבוד תחת הדרכתך. תודה שהאמנת בי ונתת לי את ההזדמנות ללמוד ממך, זה היה מסע מדהים ובלתי יסולא בפז. 
ככזה, אני אסירת תודה לא פחות לעמיתי למחקר, הדוקטורנט אריק רינברג. תודה שהיית העוגן הקבוע, שאליו אני תמיד יכולה לפנות אליו לרעיונות, לשוחח איתו וללמוד ממנו. העזרה שלך עשתה את כל ההבדל.
לחברי קבוצת המחקר של פרופסור עידית: גל אסא, עודד נאור, שיר כהן ורמי פאח׳ורי –
הזמן, הנוכחות, הרעיונות והתמיכה המתמדת שלכם היו הכרחיים, תודה לכם על כך.
ברצוני להודות למשפחתי ולחבריי על האהבה והתמיכה המתמשכת. במיוחד, אני רוצה להודות לבעלי שגיא על התמיכה האינסופית, האהבה והעידוד להשלמת המסע האקדמי שלי. תודה שתמיד גילית עניין בעבודה שלי ועזרת לי להפיק את המיטב.
תודה חייבת ללכת גם להוריי על כל האהבה והעזרה, על שתמיד האמינו בי והיו שם בשבילי.

}
